Tropical Mathematics and the Lambda-Calculus

### 2023-05-18 TLLA23/BarbaPisto.tex
commit

The study of matrices with values over the tropical semiring is a special case of the
\emph{weighted relational semantics} \cite{Manzo2013}, a well-studied semantics of the $\lambda$-calculus and linear logic:
for a fixed \emph{continuous} semi-ring $Q$, take the category $\QREL$ whose objects are sets and $\HOM \QREL X Y=Q^{X\times Y}$ (set-indexed matrices with coefficients in $Q$). %(\cite{Manzo2013} would call it $Q^\Pi$).
As expected, $Q^X$ is a $Q$-module and we can identify $\HOM{\QREL}{X}{Y}$ with the set of linear maps from $Q^X$ to $Q^Y$.
Taking $Q:=\Lawv$ we obtain the \emph{tropical weighted relational model} $\LREL$.
%, where $\Lawv$ is the already introduced Lawvere quantale, seen as the idempotent complete semiring $(\BB R_{\geq0}\cup\set{\infty},\inf,\infty,+,0)$.
%The category $\LREL$ is well-defined because $\Lawv$ is a continuous semiring (w.r.t.\ its quantale order $\preceq$.
%This amounts to check that $\min$ and $+$ commute with the $\inf$ (as operations on $\BB R_{\geq0}\cup\set{\infty}$, which is immediate), and that $(\Lawv,\preceq)$ is a cpo with $\infty$ as bottom element (which is immediate since in $\Lawv$ we have $\vee = \inf$) .
Remark that the composition in $\LREL$ %is the tropicalisation of the one defining it in $\QREL$, i.e.
reads as \ $(s\circ t)_{a,c}:=\inf_{b\in Y}\set{s_{b,c}+t_{a,b}}$;
similarly, linear maps $f:\Lawv^X\to \Lawv^Y$ are of shape $f(x)_b=\inf_{a\in X} \set{x_a+\matr f_{a,b}}$, for some matrix $\matr f\in Q^{X\times Y}$ and are precisely those induced by $\matr f\in\HOM \LREL X Y$. We call them \emph{tropical linear}.
%\end{remark}
By applying known results (taken from \cite{Manzo2013},  {\color{red}CITARE LEMAYYYY}), one obtains that $\LREL$ gives rise to denotational models of several variants of the $\lam$-calculus:
first, $\LREL$ is a SMCC, i.e.\ a model of the linear $\STLC$.
Also, the coKleisli $\LREL_!$ is CCC, i.e.\ a model of $\STLC$, where $!$ is the usual multiset comonad (so $!X$ is the set of finite multisets on $X$).
Here, the coKleisli composition of $s\in\Lawv^{!Y\times Z}$ and $t\in\Lawv^{!X\times Y}$ is the matrix $s\circ_! t\in\Lawv^{!X\times Z}$ given by
$(s\circ_! t)_{\mu,c}:=
 \inf\limits_{n\in\N, b_1\dots,b_n\in Y, \mu = \mu_1+\cdots +\mu_n}
 \left\{s_{[b_1,\dots,b_n],c} + \sum_{i=1}^n t_{\mu_i,b_i}\right\}$.
 As usual, a matrix $t\in\HOM{\LREL_!}{X}{Y}$ yields a \emph{linear} map $\Lawv^{!X}\to\Lawv^Y$. However, we can also ``express it in the base $X$'' and see it as a \emph{non-linear} map $t^!:\Lawv^X\to\Lawv^Y$, by setting 
 $t^!(x):=t\circ_! x$.
%  (we are identifying $\Lawv^X$ with the set $\HOM{\LREL_!}{\emptyset}{X}$ of the \emph{points} of $X$).
% This is the notion of \emph{non-linear} map generated by the CCC-structure of $\LREL_!$.
 Concretely, we have $t^!(x)_b=\inf_{\mu\in !X} \set{\mu x+ t_{\mu,b}}$
 where $\mu x:=\sum_{a\in X} \mu(a)x_a$.
 These functions correspond then to generalised tLs, i.e.\ with possibly infinitely many variables (as many as the elements of $X$) and for $X=Y=\set{*}$, we get usual tLs of one variable.
Instead, if the support $\set{\mu\in!X\mid\matr f_{\mu,b}\neq\infty}$ of $f$ is \emph{finite}, we get %maps of shape $f(x)_b:=\min_{\mu\in\C F} \set{\mu x+ t_{\mu,b}}$, i.e.\ 
tropical polynomials in possibly infinitely many variables.
Furthermore, $!$ can be decomposed into a family of \emph{graded} exponentials $(!_n)_{n\in\N}$ turning $(\LREL,(!_n)_{n\in\N})$ in a model for a $\mathit{BLL}$-style simply typed $\lambda$-calculus, that we call $\BSTLC$ (see e.g.~\cite{Brunel2014}).

@@

Let us now consider metric properties. First, it can be seen that all tropical \emph{linear} functions $f: \Lawv^{X}\to \Lawv^{Y}$ are non-expansive. 
%\begin{proof}[Proof sketch]
%Using the fact that $f(\B x)_{b}= \inf_{a\in X}\matr f_{a,b}+\B x_{a}$,
%the problem reduces to checking that $|(\matr f_{a,b}-\B x_{a})- (\matr f_{a,b}-\B y_{a})| = |\B x_{a}-\B y_{a}|\leq \| \B x-\B y\|_{\infty}$.\end{proof}
This result shows that, in analogy with that happens in usual metric semantics, linear programs are interpreted by non-expansive functions. 
%\begin{proof}
%Using $f(\B x)_{b}= \inf_{a\in X}\matr f_{a,b}+\B x_{a}$,
%first observe that $|(\matr f_{a,b}-\B x_{a})- (\matr f_{a,b}-\B y_{a})| = |\B x_{a}-\B y_{a}|\leq \| \B x-\B y\|_{\infty}$; we now have
%$|f(\B x)_{b}-f(\B y)_{b}| \leq |(\inf_{a\in X}\matr f_{a,b}-\B x_{a})-(\inf_{a\in X}\matr f_{a,b}-\B y_{a})| \leq
%\sup_{a\in X}|(\matr f_{a,b}-\B x_{a})- (\matr f_{a,b}-\B y_{a})|\leq 
% \| \B x-\B y\|_{\infty}$.
%\end{proof}
More generally, linear maps with bounded exponentials yield Lipschitz maps:
\begin{proposition}\label{prop:boundedlip}
If a tLs $f: \Lawv^{X}\to \Lawv^{Y}$ arises from a bounded matrix $\matr f:!_{n}X\times Y\to \Lawv$, then $f$ is $n$-Lipschitz-continuous.
\end{proposition}
%\begin{proof}[Proof sketch]
%It follows from the above lines about tropical linear functions, plus the remark that, for all $x\in \Lawv^{X}$, $\norm{ !_{n} x-!_{n} y}_{\infty}\leq n\cdot \norm{ x- y}_{\infty}$, where $!_{n} x$ is the restriction of $! x$ to finite multisets of degree $\leq n$.%
%%Using the fact that $f(\B x)_{b}=\inf_{\mu\in \C M_{\leq n}(X)}\{ \matr f_{\mu,b}+ \mu (!_{n}\B x) \}$, where $!_{n}\B x\in \Lawv^{\C M_{\leq n}(X)}$ is given by 
%%$(!_{n}\B x)_{[a_{1},\dots, a_{k}]}=\sum_{i=1}^{k}\B x_{a_{i}}$, 
%%it suffices to check that $\| (!_{n}\B x)-(!_{n}\B y)\|_{\infty}\leq n\cdot \| \B x-\B y\|_{\infty}$ and apply Proposition \ref{prop:troplinear}.
%\end{proof}
This result is perfectly analogous to what happens in the metric models recalled in the introduction.%, the bounded exponentials $!_{n}$ playing the role of the re-scaling trick.
It also entails that any tropical polynomial $\varphi:\Lawv^{X}\to\Lawv$ is $\mathrm{deg}(\varphi)$-Lipschitz continuous.

@@

Our contribution was, first, to observe that the SMCC structure of $\LREL$ lifts to that of the more general category $\Mod\equiv\GMet$; secondly, to define an exponential $!$ in $\Mod\equiv\GMet$, using a well-known recipe from \cite{Mellies2018, Manzo2013} based on the construction of symmetric algebras, and show that in this way also the exponential and differential structure of $\LREL_{!}$ lift to this more general setting:
%
%This is interesting for us because it suggests to transport the metric point of view of $\GMet$ onto our modules in $\Mod$.
%Actually, we can even show that:
\begin{theorem}\label{thm:no-base}
$\Mod_{!}$ or, equivalently, $\GMet_{!}$, can be equipped with a differential $D$ making it a $CC\partial C$.
\end{theorem}
%In order to ,, consisting in defining a symmetric algebra %$\Sym_{n}(M)$
%as the equalizer of all permutative actions on $n$-tensors $M\otimes \dots \otimes M$, and then define $!$ as the product of the symmetric algebras.

>>>>

The study of matrices with values over the tropical semiring is a special case of the
\emph{weighted relational semantics} \cite{Manzo2013}, a well-studied semantics of the $\lambda$-calculus and linear logic:
for a fixed \emph{continuous} semi-ring $Q$, take the category $\QREL$ whose objects are sets and $\HOM \QREL X Y=Q^{X\times Y}$ (set-indexed matrices with coefficients in $Q$). %(\cite{Manzo2013} would call it $Q^\Pi$).
As expected, $Q^X$ is a $Q$-module and we can identify $\HOM{\QREL}{X}{Y}$ with the set of linear maps from $Q^X$ to $Q^Y$.
Taking $Q:=\Lawv$ we obtain the \emph{tropical weighted relational model} $\LREL$.
%, where $\Lawv$ is the already introduced Lawvere quantale, seen as the idempotent complete semiring $(\BB R_{\geq0}\cup\set{\infty},\inf,\infty,+,0)$.
%The category $\LREL$ is well-defined because $\Lawv$ is a continuous semiring (w.r.t.\ its quantale order $\preceq$.
%This amounts to check that $\min$ and $+$ commute with the $\inf$ (as operations on $\BB R_{\geq0}\cup\set{\infty}$, which is immediate), and that $(\Lawv,\preceq)$ is a cpo with $\infty$ as bottom element (which is immediate since in $\Lawv$ we have $\vee = \inf$) .
Remark that the composition in $\LREL$ %is the tropicalisation of the one defining it in $\QREL$, i.e.
reads as \ $(s\circ t)_{a,c}:=\inf_{b\in Y}\set{s_{b,c}+t_{a,b}}$;
similarly, linear maps $f:\Lawv^X\to \Lawv^Y$ are of shape $f(x)_b=\inf_{a\in X} \set{x_a+\matr f_{a,b}}$, for some matrix $\matr f\in Q^{X\times Y}$ and are precisely those induced by $\matr f\in\HOM \LREL X Y$. We call them \emph{tropical linear}.
%\end{remark}
By applying known results (taken from \cite{Manzo2013}, \cite{Laird2020}, \cite{LemayCALCO2021}), one obtains that $\LREL$ gives rise to denotational models of several variants of the $\lam$-calculus:
first, $\LREL$ is a SMCC, i.e.\ a model of the linear $\STLC$.
Also, the coKleisli $\LREL_!$ is CCC, i.e.\ a model of $\STLC$, where $!$ is the usual multiset comonad (so $!X$ is the set of finite multisets on $X$).
Here, the coKleisli composition of $s\in\Lawv^{!Y\times Z}$ and $t\in\Lawv^{!X\times Y}$ is the matrix $s\circ_! t\in\Lawv^{!X\times Z}$ given by
$(s\circ_! t)_{\mu,c}:=
 \inf\limits_{n\in\N, b_1\dots,b_n\in Y, \mu = \mu_1+\cdots +\mu_n}
 \left\{s_{[b_1,\dots,b_n],c} + \sum_{i=1}^n t_{\mu_i,b_i}\right\}$.
 As usual, a matrix $t\in\HOM{\LREL_!}{X}{Y}$ yields a \emph{linear} map $\Lawv^{!X}\to\Lawv^Y$. However, we can also ``express it in the base $X$'' and see it as a \emph{non-linear} map $t^!:\Lawv^X\to\Lawv^Y$, by setting 
 $t^!(x):=t\circ_! x$.
%  (we are identifying $\Lawv^X$ with the set $\HOM{\LREL_!}{\emptyset}{X}$ of the \emph{points} of $X$).
% This is the notion of \emph{non-linear} map generated by the CCC-structure of $\LREL_!$.
 Concretely, we have $t^!(x)_b=\inf_{\mu\in !X} \set{\mu x+ t_{\mu,b}}$
 where $\mu x:=\sum_{a\in X} \mu(a)x_a$.
 These functions correspond then to generalised tLs, i.e.\ with possibly infinitely many variables (as many as the elements of $X$) and for $X=Y=\set{*}$, we get usual tLs of one variable.
Instead, if the support $\set{\mu\in!X\mid\matr f_{\mu,b}\neq\infty}$ of $f$ is \emph{finite}, we get %maps of shape $f(x)_b:=\min_{\mu\in\C F} \set{\mu x+ t_{\mu,b}}$, i.e.\ 
tropical polynomials in possibly infinitely many variables.
Furthermore, $!$ can be decomposed into a family of \emph{graded} exponentials $(!_n)_{n\in\N}$ turning $(\LREL,(!_n)_{n\in\N})$ in a model for a $\mathit{BLL}$-style simply typed $\lambda$-calculus, that we call $\BSTLC$ (see e.g.~\cite{Brunel2014}).

@@

Let us now consider metric properties. First, it can be seen that all tropical \emph{linear} functions $f: \Lawv^{X}\to \Lawv^{Y}$ are non-expansive. 
%\begin{proof}[Proof sketch]
%Using the fact that $f(\B x)_{b}= \inf_{a\in X}\matr f_{a,b}+\B x_{a}$,
%the problem reduces to checking that $|(\matr f_{a,b}-\B x_{a})- (\matr f_{a,b}-\B y_{a})| = |\B x_{a}-\B y_{a}|\leq \| \B x-\B y\|_{\infty}$.\end{proof}
This result shows that, in analogy with that happens in usual metric semantics, linear programs are interpreted by non-expansive functions. 
%\begin{proof}
%Using $f(\B x)_{b}= \inf_{a\in X}\matr f_{a,b}+\B x_{a}$,
%first observe that $|(\matr f_{a,b}-\B x_{a})- (\matr f_{a,b}-\B y_{a})| = |\B x_{a}-\B y_{a}|\leq \| \B x-\B y\|_{\infty}$; we now have
%$|f(\B x)_{b}-f(\B y)_{b}| \leq |(\inf_{a\in X}\matr f_{a,b}-\B x_{a})-(\inf_{a\in X}\matr f_{a,b}-\B y_{a})| \leq
%\sup_{a\in X}|(\matr f_{a,b}-\B x_{a})- (\matr f_{a,b}-\B y_{a})|\leq 
% \| \B x-\B y\|_{\infty}$.
%\end{proof}
More generally, linear maps with bounded exponentials yield Lipschitz maps:
\begin{proposition}\label{prop:boundedlip}
If a tLs $f: \Lawv^{X}\to \Lawv^{Y}$ arises from a bounded matrix $\matr f:!_{n}X\times Y\to \Lawv$, then $f$ is $n$-Lipschitz-continuous.
\end{proposition}
%\begin{proof}[Proof sketch]
%It follows from the above lines about tropical linear functions, plus the remark that, for all $x\in \Lawv^{X}$, $\norm{ !_{n} x-!_{n} y}_{\infty}\leq n\cdot \norm{ x- y}_{\infty}$, where $!_{n} x$ is the restriction of $! x$ to finite multisets of degree $\leq n$.%
%%Using the fact that $f(\B x)_{b}=\inf_{\mu\in \C M_{\leq n}(X)}\{ \matr f_{\mu,b}+ \mu (!_{n}\B x) \}$, where $!_{n}\B x\in \Lawv^{\C M_{\leq n}(X)}$ is given by 
%%$(!_{n}\B x)_{[a_{1},\dots, a_{k}]}=\sum_{i=1}^{k}\B x_{a_{i}}$, 
%%it suffices to check that $\| (!_{n}\B x)-(!_{n}\B y)\|_{\infty}\leq n\cdot \| \B x-\B y\|_{\infty}$ and apply Proposition \ref{prop:troplinear}.
%\end{proof}
This result is perfectly analogous to what happens in the metric models recalled in the introduction. %, the bounded exponentials $!_{n}$ playing the role of the re-scaling trick.
It also entails that any tropical polynomial $\varphi:\Lawv^{X}\to\Lawv$ is $\mathrm{deg}(\varphi)$-Lipschitz continuous.

@@

At this point, it can be shown that the SMCC structure of $\LREL$ lifts in a natural way to that of the more general category $\Mod\equiv\GMet$; secondly, an exponential $!$ in $\Mod\equiv\GMet$, can be defined via a well-known recipe  based on the construction of symmetric algebras \cite{Mellies2018, Manzo2013, Laird2020} .
Since $!$ makes $\GMet$ a Lafont category, one can either do a few calculations or apply general theorems from the literature (e.g.~\cite{LemayCALCO2021}) to obtain the following result, which lifts also the exponential and differential structure of $\LREL_{!}$ to this more general setting:
%
%This is interesting for us because it suggests to transport the metric point of view of $\GMet$ onto our modules in $\Mod$.
%Actually, we can even show that:
\begin{theorem}\label{thm:no-base}
$\Mod_{!}$ or, equivalently, $\GMet_{!}$, can be equipped with a differential $D$ making it a $CC\partial C$. 
\end{theorem}
%In order to ,, consisting in defining a symmetric algebra %$\Sym_{n}(M)$
%as the equalizer of all permutative actions on $n$-tensors $M\otimes \dots \otimes M$, and then define $!$ as the product of the symmetric algebras.
